---
title: "In-class Exercise 2"
author: "Arya Siahaan"
date: "April 20, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Getting Started**

## Loading R packages

```{r}
pacman::p_load(tidyverse, ggdist, ggridges,
               colorspace, ggthemes) 
```

# Importing the data

```{r}
exam_df <- read_csv("data/Exam_data.csv")
```

## **Visualising Distribution**

## Histogram

## **Visualising Distribution**

## Probability density plot

::: panel-tabset
## The task

```{r}
#| echo: false
ggplot(exam_df,
       aes(x = ENGLISH)) +
  geom_density(
    color = "#1696d2",
    adjust = .65,
    alpha = .6
  )
```

## The code

```{r}
#| eval: false
ggplot(exam_df,
       aes(x = ENGLISH)) +
  geom_density(
    color = "#1696d2",
    adjust = .65,
    alpha = .6
  )
```
:::

median_eng \<- median(exam_df$ENGLISH) mean_eng <- mean(exam_df$ENGLISH) std_eng \<\_ sd(exam_df\$ENGLISH)

ggplot(exam_df, aes(x = ENGLISH)) geom_density( color = "#1696d2", adjust = .65, alpha = .6) + stat_function( fun = dnorm, args = list(mean = mean_eng, sd = std_eng), col = "grey30", size = 0.8) + geom_vline( aes(xintercept = mean_eng), colour="4d5887", linewidth = 0.6, linetype = "dashed") + annotate(geom = "text", x = mean_eng - 8 y = 0.04, label = paste0("Mean ENGLISH: ", round((mean_eng), 2)), ) ) )
